---
title: "Exploratory_data_analysis"
author: "Charles Anchang"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: cosmo
    highlight: monochrome
    toc: yes
    toc_float: no
    toc_depth: 6
    code_folding: hide
  github_document:
    theme: cosmo
    highlight: monochrome
    toc: yes
    toc_float: no
    toc_depth: 6
    code_folding: hide
  word_document:
    toc: yes
    toc_depth: '6'
  pdf_document:
    toc: yes
    toc_depth: '6'
editor_options:
  chunk_output_type: console
---

```{css, echo=FALSE}
<style>
body, h1, h2, h3, h4 {
    font-family: "Bookman", serif;
}

body {
    color: #333333;
}
a, a:hover {
    color: red;
}
pre {
    font-size: 10px;
}
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## **Introduction**

RNA sequencing is a technique whereby RNA transcripts are converted to complementary DNA (cDNA) which is then sequenced to generate reads and the reads and mapped to a reference genome to generate counts which can be analysed statistically. But the statistical approaches differ than those used for microarrays because the RNA seq data consists of counts´and not intensities. Counts are modeled using the Poisson or negative binomial distribution. These techniques are implemented in the **EdgeR** and **DESeq2** packages which are some of the most widely used RNA seq data analysis. Major complications with RNA seq data analysis are the need to account for the effects of splicing and to account for isoformy. In this analysis, I will not touch details on this topic. However So, while we have approaches like edgeR, DESeq, and **DEXSeq** that basically count each gene or each exon and quantifies the amount. We also have techniques that try to determine from the reads what transcripts are present. Some of these actually try to do it from scratch, without a reference genome.
Examples of these methods are **Trinity Oases**, **Cufflinks**, and **Scripture**.

I will however look at basic statistical approaches commonly used in RNA seq analysis.

## **Load libraries**

```{r}
library(dplyr) # wrangling
```


## **Transcriptome alignment assessment**

At a previous step not detailed in this analysis, transcript- and gene-level expression was calculated by probabilistic alignment using the RSEM software. The two result are saved in the **raw_data** folder.

```{r}
list.files("raw_data")
genes <- read.table("raw_data/SRR1039508.genes.results",
                    header = TRUE, stringsAsFactors = TRUE)

head(genes) %>%
  knitr::kable()
isoforms <- read.table("raw_data/SRR1039508.isoforms.results",
                       header = TRUE, stringsAsFactors = TRUE)
head(isoforms) %>%
  knitr::kable()
```

Both genes and isoforms have a column names the **FPKM(fragments per kilobase of sequence per million mapped reads) **.

I will confirm with the *split()* and *sapply()* functions that the FPKM column in genes is the sum of the FPKM colum in isoforms.

```{r}
fpkm.per.gene <- split(isoforms$FPKM, isoforms$gene_id)
head(sapply(fpkm.per.gene, sum))
head(genes$FPKM)
```

Alternatively

```{r}
isoforms %>% group_by(gene_id) %>% summarize(sum = sum(FPKM)) 
```

I will obtain FPKM values from the genes values and log transform it and make a histogram. But before that, I will start by removing all genes with FPKM values of 0. The log transformation will be $\log10(x + 1)$.

```{r}
genes2 <- genes[genes$FPKM > 0, ]
genes2$gene_id <- droplevels(genes2$gene_id)
isoforms2 <- isoforms[isoforms$gene_id %in% genes2$gene_id, ]
isoforms2$gene_id <- droplevels(isoforms2$gene_id)
```

To verify that the gene_id column in genes2 is equal to the levels of the gene_id column in isoforms2.

```{r}
stopifnot(all(genes2$gene_id == levels(isoforms2$gene_id)))
```

Plot of effective length vs expected count:

```{r}
head(genes2)
plot(log(genes2$effective_length),
     log(genes2$expected_count))
```

**Make a histogram of the FPKM in genes2**

```{r fig.align='center'}
genes2_fpkm <- log10(genes2$FPKM)
genes2_fpkm_pos <- genes2_fpkm[genes2_fpkm > 0]
median(genes2$FPKM)
hist(log10(genes2$FPKM))
```

**With isoforms2 , calculate the maximum IsoPct for each gene**

```{r}
head(isoforms2)

max.iso <- isoforms2 %>%
              group_by(gene_id) %>%
              summarize(max_isopc = max(IsoPct))
hist(max.iso$max_isopc)
mean(max.iso$max_isopc > 95)
```

**Is there a relationship between total expression and the maximum IsoPct for each gene?**

```{r fig.align='center'}
plot(max.iso$max_isopc,
     log(max.iso$max_isopc))
boxplot(split(log10(genes2$FPKM), 
              cut(max.iso$max_isopc, 5)), xlab="max.iso", ylab="log10 FPKM")
```

**Calculate the number of isoforms per gene, and plot the maximum IsoPct against the number of isoforms:**

```{r fig.align='center'}
num.iso <- as.numeric(table(isoforms2$gene_id))
plot(num.iso, max.iso$max_isopc)

barplot(table(num.iso))
barplot(table(num.iso[num.iso < 15]))
```

**TPM (transcripts per million)**, is a linear scaling of the FPKM, such that we would expect a gene with 1 TPM to have one molecule in a population of one million mRNAs.

With genes2, plot the TPM over the FPKM.

Confirm that TPM is equal to: (FPKM / sum(FPKM)) * 1e6

